---
title: "Evaluating Bias in Baseball Statistics Using Umpire Fixed Effects"
author: "Evan Flack"
date: "7/9/2020"
output: html_document
---

### Introduction

Baseball is famously, and often comically, a game of statistics. Watch any broadcast and you will likely hear some version of the following: "this is the first time a player from Texas has hit a three-run homerun in the rain on a Tuesdy since 1981." Obsurce facts aside, teams, journalists, and fans alike are increasibgly using statistics to better understand the game. Most often, statistics are used for prediction: scouts use college statistics to predict how well a player will do in the professionals, TV broadcasts show conditional averages (e.g. batting average with runners in scoring position) to give the viewer a sense of the probability the batter will get a hit. 

Sometimes, these predictions are mistakenly interpreted to have causal implications. Take for example, a player's batting average conditional on the pitch count. It is conventional wisdom that "batters' counts" (e.g. 3-1) are easier to hit in than "pitcher' counts" (e.g. 1-2), because if the pitchers runs the risk of walking the batter, he must try and throw a stike, or coloquially, "give the batter something to hit." This wisodom is thought to be backed up by statistics: the MLB batting average in 3-1 counts is X, while the average in 1-2 counts is substantially lower (X). However, the count is not exogenous to the situation. It is likely the case that worse pitchers end up in batters' counts, and so some of the difference in the conditional averages above reflects overall pitcher ability, and not the count itself. One may think this is easy to solve, pitcher ability can be measured, so we need only include measures such as earned run average (ERA) or strikeouts to eliminate potnetial omitted vaiables bias (OMB). Ask anyone in baseball however, and they will tell you that pitcher ability varies widely from game to game. For example Dallas Braden threw a perfect game in 2010, a year whre he a good but not great ERA of 3.50. I explore variablility in pitcher performance more in-depth, but this example should alert you to the fact that even after including overall measures of pitcher performance, there are likely relevant omitted variables that we are unable to account for.

To adress OMB in the case presented above, and potentially others, I propose a new method that uses umpire fixed effects to generate exogenous variation in the count that batters face. The methodology is closely based on the judge fixed effects designs widely used by economists in crime research (for an overview see X). The basic presmise is that umpires (judges) are randomly assigned to games (cases), and that each umpire (judge) varies in their harshness towards batters (defendants), and therefore this harshness is randonly assigned. In the judge fixed effects case, harshness is often defined as the probability of putting a defendant in jail, while in the umpire case one could say that judge who call more strikes (especially on close pitches) are harsher towards batters. The harshness of the judge is used as an instument for say, a prison sentence, and the efects of the prsion sentence are evaluated in a second stage. There are of course many nuanced differences between the cases of umpires and judges, which I hope to explore in future work. 

The purpose of this essay is to (1) show that umpires are (as if) randomly assigned to teams/games, and show that there is substantial heterogeneity in their harshness toward batters. I first briefly explain the gameplay of baseball focussing on the home plate umpire's role. I then estimate umpire fixed effects and show that there is substantial heterogentiy in the propensity to call strikes, which greatly increases when I subset to a sample of marginal pitches. Finally, I estimate a first stage of a strike call on the umpire's average propensity to call a strike, leaving out calls from the same game when calculating this average. I leave the estimation of a second stage, using an outcome such as getting on base, to later work, and would gratefully accept suggestions/collaboration in using this design to evaluate outcomes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
library(grid)
```

### Strike Zone
The left panel of figue X shows the standard strike zone (rectangle), and the loction of 100 randomly selected pitches from the 2016 that were taken by batters. On average, umpires do a god jib of making the right call: most pitches inside the strike zone are called stikes (triangles) and most outside are balls (circles). There is considerably ambiguity near the edges (margins) of the strike zone howver, with some pitches just inside being called balls, and some that land just outsie called strikes. The right panel of figure X highlights these marginal areas, by extending the strike zone both slightly inward (.05 units) and more substantially outward (0.15 unite), to define an area where roughly half of the pitches are called balls and strikes respecitively. If a pitch lands in the area defined by the dashed lines, one could think of the call of ball/strike as the outcome of a coin flip.

```{r fig.width=8, fig.height=5,echo=FALSE, fig.align="center"}
img <- readPNG("strike_zone.png")
grid.raster(img)
```


### Model

The situation I would like to emulate is the following: imagine that every time a batter came to the plate the umpire flipped a coin, if it landed on heads the count began at 1-0, if tails, the count goes to 0-1.

$y_{1,1} - y_{0, 1} + y_{0, 1} - y_{0, 0}$

$E[hit | strike] - E[hit | ball]$


### Umpire Fixed Effects

What the coin-flip does not capture however, is that different umpires may vary systematically in ther interpretation of close pitches, i.e. not all umpires call strike 50 percent of the time for a ball on the margin. Ome umpire may have an expanded strike zone, only calling 40 percent strikes. This mirrors the case of judges, where the overall liklihood of going to jail may be a coin flip, but individual judges vary widely in their propensity to use jail as a punishment. Figure X shows estimateds umpire fixed effects for the propensity to call a strike both for all pitches (left panel) and those that I define to be on the margin (right panel). 

```{r fig.width=8, fig.height=4,echo=FALSE, fig.align="center"}
img <- readPNG("umpire_fe.png")
grid.raster(img)
```

### First Stage Estimate

|                 |Term                    |Estimate              |Observations |% Strike |
|:----------------|:-----------------------|:---------------------|:------------|:--------|
|All Pitches      |Intercept               |0.135*** (0.0231)     |360000       |0.335    |
|                 |Umpire Strike Avg (LOO) |0.597*** (0.0688)     |             |         |
|Marginal Pitches |Intercept               |0.139*** (0.024)      |139000       |0.53     |
|                 |Umpire Strike Avg (LOO) |0.737*** (0.0452)     |             |         |
|Clear Strikes    |Intercept               |0.566*** (0.0722)     |46000        |0.983    |
|                 |Umpire Strike Avg (LOO) |0.425*** (0.0734)     |             |         |
|Clear Balls      |Intercept               |0.00395*** (0.000656) |175000       |0.0115   |
|                 |Umpire Strike Avg (LOO) |0.658*** (0.0525)     |             |         |
